Pattern matching for\\ $(213,231)$-avoiding permutations

### 2015-02-05 article.tex
typos

\title{Pattern matching for\\ $(231,213)$-avoiding permutations}
\author{Both Emerite NEOU, Romeo RIZZI}
\date{}

@@

	A few particular cases of the pattern containment problem have been attacked successfully.
	The case of increasing patterns is solvable in
	$O(|\pi| \log \log |\sigma|)$ time in the RAM model \cite{Crochemore:Porat:2010},
	improving the previous 30-year bound of $O(|\pi| \log |\sigma|)$.
	(The algorithm also improves on the previous
	$O(|\pi| \log \log |\pi|)$ bound.)
	Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear time
	by stack sorting algorithms.
	Any pattern of length $4$ can be detected in $O(|\pi| \log |\pi|)$ time
	\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
	Algorithmic issues for $321$-avoiding patterns containment has been investigated in
	\cite{Guillemot:Vialette:ISAAC:2009}.
	The pattern containment problem is also solvable in
	polynomial-time for separable patterns \cite{Ibarra:1997}
	(see also \cite{Bose:Buss:Lubiw:1998}).
	Separable permutations are those permutations that contain neither
	$2413$ nor $3142$, and they are enumerated by the Schröder numbers
	(sequence A006318 in OEIS).
	To the best of our knowledge,
	separable permutations first arose in the work of
	Avis and Newborn~\cite{Avis:Newborn:1981},
	who showed that they are precisely the permutations which can be sorted by an
	arbitrary number of pop-stacks in series,
	where a pop-stack is a restricted form of stack in which any pop operation
	pops all items at once.
	(Notice that the separable permutations include as a special case the
	stack-sortable permutations, which avoid the pattern $231$.)

	There exist many generalisation of patterns that are worth considering
	in the context of algorithmic issues in pattern involvement
	(see \cite{Kitaev:book:2011} for an up-to-date survey).
	\emph{Vincular patterns}, also called 
	\emph{generalized patterns},
	resemble (classical) patterns, with the constraint that some of the letters in
	an occurrence must be consecutive.
	Of particular importance in our context,
	Bruner and Lackner \cite{Bruner:Lackner:SWAT:2012}
	proved that deciding whether a vincular pattern
	$\sigma$ of length $k$ occurs in a permutation
	$\pi \in \mathfrak{S}_n$ is $W[1]$-complete for
	parameter $k$.
	\emph{Bivincular patterns} generalize classical patterns even further 
	than vincular
	patterns. Indeed, in bivincular patterns,
	not only positions but also values
	of elements involved in a matching may be forced to be adjacent

@@

A \emph{permutation} of length $n$ is a one-to-one function from an
$n$-element set to itself.
We write permutations as words 
$\sigma = \sigma_1\sigma_2\,\ldots\,\sigma_n$, whose letters are distinct
and uasually consist of the integers $12\,\ldots\,n$.
For the sake of convenience, 
we let 
$\pi[i:j]$ stand for 
$\pi_i\pi_{i+1}\,\ldots\,\pi_j$,
$\pi[:j]$ stand for $\pi[1:j]$ and
$\pi[i:]$ stand for $\pi[i:n]$.
We let $S_n$ denote the set of all permutations of length $n$.

@@

An \emph{ascent} of a permutation $\pi \in S_n$ is any position 
$1 \leq i < n$ where the following value is bigger than the current one. 
That is, if $\pi = \pi_1\pi_2\,\ldots\,\pi_n$, then
$i$ is an ascent if $\pi_i < \pi_{i+1}$.
For example, the permutation 
$345216$7 has ascents (at positions) $1,2,5,6$.
Similarly, a \emph{descent} is a position 
$1 \leq i < n$ with $\pi_i > \pi_{i+1}$, 
so every $i$ with $1 \leq i < n$ either is an ascent or is a descent of 
$\pi$.
Let $\ustep$ and $\dstep$ denote an ascend and a descend, respectively.
The \emph{stripe} $s_\pi$ of a permutation $\pi \in S_n$ is the word
$\stripe{\pi}{1} \stripe{\pi}{2} \ldots \stripe{\pi}{n-1} \in \{\ustep,\dstep\}^{n-1}$ 
defined by 
$ \stripe{\pi}{i}= \ustep$ if $i$ an ascent in $\pi$ and
$\stripe{\pi}{i} = \dstep$ if $i$ a descent in $\pi$.
For example the stripe of the permutation 
$\pi = 12\,11\,10\,1\,2\,3,4\,9\,8\,5\,6\,7$ 
is 
$\stripew{\pi} = \dstep\dstep\dstep\ustep\ustep\ustep\ustep\dstep\dstep\ustep\ustep$.
The stripe $\stripew{\sigma}$ is a \emph{pattern} of the strip $\stripew{\pi}$
(or $\stripew{\sigma}$ \emph{occurs} of the strip $\stripew{\pi}$) if $\stripew{\sigma}$ occurs in
$\stripew{\pi}$ as a subsequence.

A \emph{bivincular pattern} (abbreviated BVP) $\sigma$
of length $k$ is a permutation in $S_k$ written in
two-line notation 
(that is the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$).
We have the following conditions on the top and bottom rows
of $\sigma$:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the letters corresponding to 
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in an occurrence of
	$\sigma$ in a permutaton must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive letters.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must begin with the leftmost 
	letter of $\pi$,
	and 
	if the bottom row of $\sigma$ begins with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must end with the rightmost 
	letter of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the letters corresponding to
	$\sigma_i, \sigma_{i+1}, \ldots, \sigma_j$ in an
	occurrence of $\sigma$ in a permutation must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined 
	letters.
	Moreover, if the top row of $\sigma$ begins with
	$^ulcorner{1}$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must begin with
	the smallest letter of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must end with
	the largest letter of $\pi$.
\end{itemize}
A bivincular pattern avoids a set of permutations $P$ if its bottom
row $\sigma_1\sigma_2\,\ldots\,\sigma$ avoids every permutation in $P$.

@@

\section{Both $\pi$ and $\sigma$ are $(231,213)$-avoiding}
\label{section:both are (231,213)-avoiding}

This section is devoted to presenting a fast algorithm for deciding
$\sigma \leq \pi$ 
in case both $\pi$ and $\sigma$ are $(231,213)$-avoiding.

@@

\textbf{Proposition  \num \thenum} Let $\pmotif$ and $\ptext$
be two permutations avoiding (231,213), 
we can decide in linear time if $\pmotif$ 
appear in $\ptext$.\\

@@

%		\subsection{With Motif Avoiding (231,213)}	
%		
%			\textbf{Proposition \num} Let $\pmotif$, $\ptext$ 
%			be two permutation,
%			with  $\pmotif$ a permutation avoiding (231,213), 
%			we can decide in $O(|\pmotif|*|\ptext|^3)$ time
%			and $O(|\ptext|^3) space$
%			if $\pmotif$ 
%			appear in $\ptext$.\\	
%
%			\textbf{Proof}
%			We consider the following problem :\\
%			
%			Given a motif $\pmotif$ avoiding $(231,213)$,
%			a text $\ptext$, $i$,$j$ , $i<j$.\\
%			
%			$\mathbb{P}_{\pmotif,\ptext,min,max}(i,j)= \left\{ 
%				\begin{array}{l l}
%					true & \text{If the sufixe of $\pmotif$ starting at $i$}\\
%						& \text{is a pattern of the suffix of $\ptext$} \\
%						& \text{with every element in $[min,max]$}\\
%		
%					false & otherwise\\
%				\end{array} \right.$\\
%		
%			$\mathbb{P}_{\pmotif,\ptext,min,max}$ is closed under induction. 
%			it can be
%			solved by means of the following relations:\\
%			
%			
%			\textbf{BASE :} \\
%			$\mathbb{P}_{\pmotif,\ptext,min,max}(|\pmotif|,j)= \left\{ 
%					\begin{array}{l l}
%						true & \text{if $min < \ptext(j) < max$
%						}\\
%						false & otherwise\\
%					\end{array} \right. $\\	
%		
%			\textbf{STEP :} \\
%			$\mathbb{P}_{\pmotif,\ptext,min,max}(i,j)= \left\{ 
%					\begin{array}{l l}
%						False & \text{if $\ptext(j)<min$ or $max<\ptext(j)$}\\
%						\bigcup_{j<k}\mathbb{P}_{\pmotif,\ptext,t(j),max}(i+1,k) 
%							& \text{if $\\mathcal{S}(\pmotif,i)=\ustep$
%							}\\
%						\bigcup_{j<k}\mathbb{P}_{\pmotif,\ptext,min,t(j)}(i+1,k) 
%							& \text{if $\\mathcal{S}(\pmotif,i)=\dstep$
%							}\\
%					\end{array} \right. $\\
%			At each step, 
%			we are match the current element of $\pmotif$ ($\pmotif(i)$)
%			with the current element of $\ptext$ ($\ptext(j)$), if possible.
%			then we are match the rest of $\pmotif$ with 
%			every suffixes $\ptext$ starting after $j$.
%			Clearly there is a match if and only if 	$\bigcup_{0<k<|\ptext|,0<min<max<|\ptext|}\mathbb{P}_{\pmotif,\ptext,min,t(j)}(0,k)$
%			is true.   
%			This implies a $O(|\pmotif|*|\ptext|^3)$ time
%			and $O(|\ptext|^3) space$ algorithm.\\

@@

This section focus on the pattern matching problem when only the motif avoid $(231,213)$.

@@

This section gives an algorithm to detect bivincular permutation in polynomial time in the case where the motif avoid $(231,213)$.

>>>>

\title{Pattern matching for\\ $(213,231)$-avoiding permutations}
\author{Both Emerite NEOU, Romeo RIZZI}
\date{}

@@

	A few particular cases of the pattern containment problem have been attacked successfully.
	The case of increasing patterns is solvable in
	$O(|\pi| \log \log |\sigma|)$ time in the RAM model \cite{Crochemore:Porat:2010},
	improving the previous 30-year bound of $O(|\pi| \log |\sigma|)$.
	(The algorithm also improves on the previous
	$O(|\pi| \log \log |\pi|)$ bound.)
	Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
	by stack sorting algorithms.
	Any pattern of length $4$ can be detected in $O(|\pi| \log |\pi|)$ time
	\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
	Algorithmic issues for $321$-avoiding patterns containment has been investigated in
	\cite{Guillemot:Vialette:ISAAC:2009}.
	The pattern containment problem is also solvable in
	polynomial-time for separable patterns \cite{Ibarra:1997}
	(see also \cite{Bose:Buss:Lubiw:1998}).
	Separable permutations are those permutations that contain neither
	$2413$ nor $3142$, and they are enumerated by the Schröder numbers
	(sequence A006318 in OEIS).
	To the best of our knowledge,
	separable permutations first arose in the work of
	Avis and Newborn~\cite{Avis:Newborn:1981},
	who showed that they are precisely the permutations which can be sorted by an
	arbitrary number of pop-stacks in series,
	where a pop-stack is a restricted form of stack in which any pop operation
	pops all items at once.
	(Notice that the separable permutations include as a special case the
	stack-sortable permutations, which avoid the pattern $231$.)

	There exist many generalisation of patterns that are worth considering
	in the context of algorithmic issues in pattern involvement
	(see \cite{Kitaev:book:2011} for an up-to-date survey).
	\emph{Vincular patterns}, also called 
	\emph{generalized patterns},
	resemble (classical) patterns, with the constraint that some of the letters in
	an occurrence must be consecutive.
	Of particular importance in our context,
	Bruner and Lackner \cite{Bruner:Lackner:SWAT:2012}
	proved that deciding whether a vincular pattern
	$\sigma$ of length $k$ occurs in a longer permutation
	$\pi$ is $W[1]$-complete for
	parameter $k$.
	\emph{Bivincular patterns} generalize classical patterns even further 
	than vincular
	patterns. Indeed, in bivincular patterns,
	not only positions but also values
	of elements involved in a matching may be forced to be adjacent

@@

A \emph{permutation} of length $n$ is a one-to-one function from an
$n$-element set to itself.
We write permutations as words 
$\sigma = \sigma_1\sigma_2\,\ldots\,\sigma_n$, whose letters are distinct
and uasually consist of the integers $12\,\ldots\,n$.
We let 
$\pi[i]$ stands for $\pi_i$.
For the sake of convenience, we let
$\pi[i:j]$ stand for 
$\pi_i\pi_{i+1}\,\ldots\,\pi_j$,
$\pi[:j]$ stand for $\pi[1:j]$ and
$\pi[i:]$ stand for $\pi[i:n]$.
As usual, we let $S_n$ denote the set of all permutations of length $n$.

@@

An \emph{ascent} of a permutation $\pi \in S_n$ is any position 
$1 \leq i < n$ where the following value is bigger than the current one. 
That is, if $\pi = \pi_1\pi_2\,\ldots\,\pi_n$, then
$i$ is an ascent if $\pi_i < \pi_{i+1}$.
For example, the permutation 
$345216$7 has ascents (at positions) $1,2,5,6$.
Similarly, a \emph{descent} is a position 
$1 \leq i < n$ with $\pi_i > \pi_{i+1}$, 
so every $i$ with $1 \leq i < n$ either is an ascent or is a descent of 
$\pi$.
Let $\ustep$ and $\dstep$ denote an ascend and a descend, respectively.
The \emph{stripe} $s_\pi$ of a permutation $\pi \in S_n$ is the word
$\stripe{\pi}{1} \stripe{\pi}{2} \ldots \stripe{\pi}{n-1} \in \{\ustep,\dstep\}^{n-1}$ 
defined by 
$ \stripe{\pi}{i}= \ustep$ if $i$ an ascent in $\pi$ and
$\stripe{\pi}{i} = \dstep$ if $i$ a descent in $\pi$.
For example the stripe of the permutation 
$\pi = 981234765$
is $\stripew{\pi} = \dstep\dstep\ustep\ustep\ustep\ustep\dstep\dstep$.
The stripe $\stripew{\sigma}$ is a \emph{pattern} of the strip $\stripew{\pi}$
(or $\stripew{\sigma}$ \emph{occurs} of the strip $\stripew{\pi}$) if $\stripew{\sigma}$ occurs in
$\stripew{\pi}$ as a subsequence.

A \emph{bivincular pattern} (abbreviated BVP) $\sigma$
of length $k$ is a permutation in $S_k$ written in
two-line notation 
(that is the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$).
We have the following conditions on the top and bottom rows
of $\sigma$:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the letters corresponding to 
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in an occurrence of
	$\sigma$ in a permutaton must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive letters.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must begin with the leftmost 
	letter of $\pi$,
	and 
	if the bottom row of $\sigma$ begins with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must end with the rightmost 
	letter of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the letters corresponding to
	$\sigma_i, \sigma_{i+1}, \ldots, \sigma_j$ in an
	occurrence of $\sigma$ in a permutation must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined 
	letters.
	Moreover, if the top row of $\sigma$ begins with
	$^\ulcorner{1}$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must begin with
	the smallest letter of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must end with
	the largest letter of $\pi$.
\end{itemize}
A bivincular pattern avoids a set of permutations $P$ if its bottom
row $\sigma_1\sigma_2\,\ldots\,\sigma$ avoids every permutation in $P$.

@@

\section{Both $\pi$ and $\sigma$ are $(213,231)$-avoiding}
\label{section:both are (231,213)-avoiding}

This section is devoted to presenting a fast algorithm for deciding
$\sigma \leq \pi$ 
in case both $\pi$ and $\sigma$ are $(213,231)$-avoiding.
We begin with an easy structure lemma.

@@

\textbf{Proposition  \num \thenum} Let $\pmotif$ and $\ptext$
be two permutations avoiding (231,213), 
we can decide in linear-time if $\pmotif$ 
appear in $\ptext$.

@@

%		\subsection{With Motif Avoiding (231,213)}	
%		
%			\textbf{Proposition \num} Let $\pmotif$, $\ptext$ 
%			be two permutation,
%			with  $\pmotif$ a permutation avoiding (231,213), 
%			we can decide in $O(|\pmotif|*|\ptext|^3)$ time
%			and $O(|\ptext|^3) space$
%			if $\pmotif$ 
%			appear in $\ptext$.\\	
%
%			\textbf{Proof}
%			We consider the following problem :\\
%			
%			Given a motif $\pmotif$ avoiding $(213,231)$,
%			a text $\ptext$, $i$,$j$ , $i<j$.\\
%			
%			$\mathbb{P}_{\pmotif,\ptext,min,max}(i,j)= \left\{ 
%				\begin{array}{l l}
%					true & \text{If the sufixe of $\pmotif$ starting at $i$}\\
%						& \text{is a pattern of the suffix of $\ptext$} \\
%						& \text{with every element in $[min,max]$}\\
%		
%					false & otherwise\\
%				\end{array} \right.$\\
%		
%			$\mathbb{P}_{\pmotif,\ptext,min,max}$ is closed under induction. 
%			it can be
%			solved by means of the following relations:\\
%			
%			
%			\textbf{BASE :} \\
%			$\mathbb{P}_{\pmotif,\ptext,min,max}(|\pmotif|,j)= \left\{ 
%					\begin{array}{l l}
%						true & \text{if $min < \ptext(j) < max$
%						}\\
%						false & otherwise\\
%					\end{array} \right. $\\	
%		
%			\textbf{STEP :} \\
%			$\mathbb{P}_{\pmotif,\ptext,min,max}(i,j)= \left\{ 
%					\begin{array}{l l}
%						False & \text{if $\ptext(j)<min$ or $max<\ptext(j)$}\\
%						\bigcup_{j<k}\mathbb{P}_{\pmotif,\ptext,t(j),max}(i+1,k) 
%							& \text{if $\\mathcal{S}(\pmotif,i)=\ustep$
%							}\\
%						\bigcup_{j<k}\mathbb{P}_{\pmotif,\ptext,min,t(j)}(i+1,k) 
%							& \text{if $\\mathcal{S}(\pmotif,i)=\dstep$
%							}\\
%					\end{array} \right. $\\
%			At each step, 
%			we are match the current element of $\pmotif$ ($\pmotif(i)$)
%			with the current element of $\ptext$ ($\ptext(j)$), if possible.
%			then we are match the rest of $\pmotif$ with 
%			every suffixes $\ptext$ starting after $j$.
%			Clearly there is a match if and only if 	$\bigcup_{0<k<|\ptext|,0<min<max<|\ptext|}\mathbb{P}_{\pmotif,\ptext,min,t(j)}(0,k)$
%			is true.   
%			This implies a $O(|\pmotif|*|\ptext|^3)$ time
%			and $O(|\ptext|^3) space$ algorithm.\\

@@

This section focus on the pattern matching problem when only the motif avoid $(213,231)$.

@@

This section gives an algorithm to detect bivincular permutation in polynomial time in the case where the motif avoid $(213,231)$.
